---
title: "Crude Oil Analysis"
author: "Kevin Bonds"
date: "10/29/2019"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The following is an analysis of crude oil data. There are many factors that come into play to influence the prices of petroleum products. An exhaustive study would take some time, but we can gain some insights in a short amount of time and prepare for deeper analysis. 

The first thing to do is to load the libraries needed

```{r message=FALSE, warning=FALSE, error=FALSE}
library(tidyverse)
library(readxl)
library(lubridate)
library(zoo)
library(knitr)
```

## Load the data

The xls file can be downloaded from: https://www.eia.gov/dnav/pet/PET_PRI_SPT_S1_M.htm. The series history can be loaded by selecting the excel icon. Let's quickly create a csv called _spot_crude.csv_ of just the first data tab. We will load all tabs, from the xls directly, in a later step. Now let's load the csv into R to start formatting. 

```{r load_csv, message=FALSE, warning=FALSE}
# Quickly read in WTI crude oil from csv
wti_crude_spot <- read_csv("spot_crude.csv")
wti_crude_spot <- wti_crude_spot %>% 
        select(1:3) 
head(wti_crude_spot)
```

## Formatting and adding MoM and YoY

The Tidyverse has some great functions. One of them is the _lag()_ function. It allows you to lag or offset some number of records that you specify. I used the default 1 for MoM and 12 for YoY (as seen below). The data is already sorted. 

I also pulled out _Month_ and _Year_ to make it a little simpler to join on later. 

```{r first_df}
# Create a clean df with MoM and YoY for just Crude Oil
wti_crude_spot <- wti_crude_spot %>% 
        mutate("Date2" = as.Date(as.yearmon(Date, "%b-%Y"), frac = 1),
               "Month" = month(Date2),
               "Year" = year(Date2),
               "MoM_crude_oil" = 
                       (`Cushing, OK WTI Spot Price FOB (Dollars per Barrel)` - 
                                lag(`Cushing, OK WTI Spot Price FOB (Dollars per Barrel)`))/ 
                       lag(`Cushing, OK WTI Spot Price FOB (Dollars per Barrel)`),
               "YoY_crude_oil" = 
                       (`Cushing, OK WTI Spot Price FOB (Dollars per Barrel)` - 
                                lag(`Cushing, OK WTI Spot Price FOB (Dollars per Barrel)`, 12))/ 
                       lag(`Cushing, OK WTI Spot Price FOB (Dollars per Barrel)`, 12))

kable(wti_crude_spot[12:17,], caption= "Table with MoM and YoY")
```

## Calculate some yearly stats

We can also quickly calcultate some yearly stats and join them back to the original table. This may come in handy as features for a model. It's quick and easy so let's go ahead and do that now.

```{r yearly_stats, message=FALSE, warning=FALSE}
# Calculate yearly stats
year_stats <- wti_crude_spot %>% 
        group_by(Year) %>% 
        summarize( "yr_mean_crude" = mean(`Cushing, OK WTI Spot Price FOB (Dollars per Barrel)`),
                   "yr_median_crude" = median(`Cushing, OK WTI Spot Price FOB (Dollars per Barrel)`))
# Join to larger dataframe
wti_crude_spot <- left_join(wti_crude_spot, year_stats, on = c("Year" = "Year"))
kable(wti_crude_spot[12:17,], caption= "Table with Yearly Stats")
```

## Loading the rest of the data from the xlsx

I want to read each tab of the xlsx file into it's own data frame to allow specifying column types--since the tabs differ in format a bit. They all have a few rows that need to be skipped so let's do that as we read them in.

```{r rest_of_data, warning=FALSE, message=FALSE}
# Read rest of data directly from xlsx file into tables
raw_data_path <- "raw_data_sheet.xlsx"
sheets <- raw_data_path %>%
        excel_sheets() %>% 
        set_names()

conv_gasoline <- read_excel(raw_data_path, sheet = sheets[3], skip = 2, col_types = c("date", "numeric", "numeric")) %>% 
        mutate("Month" = month(Date), "Year" = year(Date))

RBOB_gasoline <- read_excel(raw_data_path, sheet = sheets[4], skip = 2, col_types = c("date", "numeric")) %>% 
        mutate("Month" = month(Date), "Year" = year(Date))

heating_oil <- read_excel(raw_data_path, sheet = sheets[5], skip = 2, col_types = c("date", "numeric")) %>% 
        mutate("Month" = month(Date), "Year" = year(Date))

uls_diesel <- read_excel(raw_data_path, sheet = sheets[6], skip = 2, col_types = c("date", "numeric", "numeric", "numeric")) %>% 
        mutate("Month" = month(Date), "Year" = year(Date))

jet <- read_excel(raw_data_path, sheet = sheets[7], skip = 2, col_types = c("date", "numeric")) %>% 
        mutate("Month" = month(Date), "Year" = year(Date))
propane <- read_excel(raw_data_path, sheet = sheets[8], skip = 2, col_types = c("date", "numeric")) %>% 
        mutate("Month" = month(Date), "Year" = year(Date))

```

## Join all data together on _Year_ and _Month_

Next, let's join all tables, clean and organize the resulting table a bit.

```{r join_all, warning=FALSE, message=FALSE}
# Join conv_gasoline and heating_oil
energy_df <- left_join(wti_crude_spot, conv_gasoline[,2:5], on = c("Year" = "Year", "Month" = "Month")) %>% 
        left_join(heating_oil[,2:4], on = c("Year" = "Year", "Month" = "Month")) %>%
        left_join(uls_diesel[-1], on = c("Year" = "Year", "Month" = "Month")) %>% 
        left_join(RBOB_gasoline[-1], on = c("Year" = "Year", "Month" = "Month")) %>% 
        left_join(jet[-1], on = c("Year" = "Year", "Month" = "Month")) %>% 
        left_join(propane[-1], on = c("Year" = "Year", "Month" = "Month"))

energy_df <- energy_df %>% select("Date"= `Date2`, c(5:6, 2:3, 7:length(energy_df)))
head(energy_df)
```
## All the Variables
Now we have some data, but not much along the lines of outside influences to explain the changes in crude oil prices. One thing, I can think of, might be US oil production levels. Let's grab that data and join it in.

```{r production_levels, message=FALSE, warning=FALSE}
# Add US crude oil production form https://www.eia.gov/dnav/pet/hist/LeafHandler.ashx?n=PET&s=MCRFPUS1&f=M
US_crude_prod <- read_excel("US_crude_prod.xls", sheet = sheets[2], skip = 2, col_types = c("date", "numeric")) %>% 
        mutate("Month" = month(Date), "Year" = year(Date))

energy_df <- left_join(energy_df, US_crude_prod[-1], on = c("Year" = "Year", "Month" = "Month"))
```
## Now we can start looking at the data

Let's create some time series objects and plot them

```{r ts_plots, fig.height=10}
# Create time series objects for WTI Spot and Field Production
cop <-  ts(energy_df$`Cushing, OK WTI Spot Price FOB (Dollars per Barrel)`, start= c(1986,1), end = c(2019,8), frequency = 12)
prod_cude <-  ts(energy_df$`U.S. Field Production of Crude Oil (Thousand Barrels)`, start= c(1986,1), end = c(2019,8), frequency = 12)
par(mfrow=c(2,1))
# Plot them one on top of the other
plot(cop, type = "o", ylab = "Price", main = "Fig. 1 - Time series plot for Crude oil price")
plot(prod_cude, type = "o", ylab = "Monthly Barrels", main = "Fig. 2 - Time series plot for US Crude Oil Production")

```

Obviously we have some inflation impact to consider but, we can see there has been quite a bit of volatility even smoothing to month. We could check to see where the max price, min price, max MoM and min MoM (which would be the max decrease). Also, maybe if we look at only recent prices we can see if there is some correlation.

```{r}
# Max crude oil price
kable(energy_df[which.max(energy_df$`Cushing, OK WTI Spot Price FOB (Dollars per Barrel)`),][c(1,4)])

# Min crude oil price
kable(energy_df[which.min(energy_df$`Cushing, OK WTI Spot Price FOB (Dollars per Barrel)`),][c(1,4)])

# Max MoM
kable(energy_df[which.max(energy_df$MoM_crude_oil),][c(1,6)])

# min MoM
kable(energy_df[which.min(energy_df$MoM_crude_oil),][c(1,6)])
```

```{r}
# Max MoM
kable(energy_df[which.max(energy_df$MoM_crude_oil),][c(1,6)])
```


```{r}
# Jan-2013 to July-2019
cor(energy_df$`Cushing, OK WTI Spot Price FOB (Dollars per Barrel)`[325:403], energy_df$`U.S. Field Production of Crude Oil (Thousand Barrels)`[325:403])
# Feb-2015 to July-2019
cor(energy_df$`Cushing, OK WTI Spot Price FOB (Dollars per Barrel)`[350:403], energy_df$`U.S. Field Production of Crude Oil (Thousand Barrels)`[350:403])
# Jan-2016 to July-2019
cor(energy_df$`Cushing, OK WTI Spot Price FOB (Dollars per Barrel)`[361:403], energy_df$`U.S. Field Production of Crude Oil (Thousand Barrels)`[361:403])

```

Looks like as we take more recent data there is more correlation. This suggests that adjusting for inflation may show more correlation. 

## Extra Credit

```{r}
# How tightly correlated is Gulf Coast Spot prices vs New York Harbor?
cor(energy_df$`U.S. Gulf Coast Conventional Gasoline Regular Spot Price FOB (Dollars per Gallon)`[6:405], energy_df$`New York Harbor Conventional Gasoline Regular Spot Price FOB (Dollars per Gallon)`[6:405])
```

What about WTI vs US Gulf Coast spot?


```{r}
# How tightly correlated is Gulf Coast Spot prices vs Cushing, OK WTI?
cor(energy_df$`U.S. Gulf Coast Conventional Gasoline Regular Spot Price FOB (Dollars per Gallon)`[6:405], energy_df$`Cushing, OK WTI Spot Price FOB (Dollars per Barrel)`[6:405])
```

## Next steps

Some next steps would be to adjust our prices for inflation; try fitting a linear model using the features we have; Do some investigating around seasonality